Supporting information: Tail associations and their impact on extinction risk

### 2019-10-25 MainText.Rmd
working on methods, also some other points made

---
title: "Asymmetric tail association in extreme climatic events affects population-extinction risk"
author: "Shyamolina Ghosh, Daniel Reuman"
date: "`r format(Sys.time(), '%d %B, %Y')`"
geometry: "left=1in,right=1in,top=1in,bottom=1in"
output: 
  pdf_document:
    number_sections: no
    keep_tex: yes
    fig_caption: yes
    includes:
      in_header: head_maintext.sty

@@

<!--
Alternative titles
1) Tail associations in environmental variables and extinction risk
2) The influence of correlations in the extremes on extinction risk
DAN FAVORITE SO FAR 3) Tail associations in environmental variables and their impact on extinction risk
-->

@@

<!--ECEs have a lot of impacts, it is important to understand their impacts-->
Over the past few decades, extreme climatic events (ECEs) such as
temperature or precipitation extremes, heat-waves, and extreme wildfires
have become more common or more extreme due to anthropogenic climate change 
[@ummenhofer2017extreme]. ECEs are well known to produce severe
impacts at the population, community and ecosystem levels [@bragazza2008climatic; 
@gutschick2003extreme; @jiguet2011community; @mckechnie2009climate;
@felton2017integrating], and thus the goal of improving our understanding of
the variety of ecological impacts of ECEs is an important applied goal for 
modern ecology [@katz2005statistics; @smith2011ecological; @diez2012will; @moritz2013future; @bailey2016tackling]. Frequent extreme events 
can help alter the mean temperature or precipitation of an area, 
can be associated with changes through time in the overall variance of 
climate signals, and can directly contribute to changes in the symmetry or the 
skewness of distributions of climate signals through time 
[@hansen2012perception]. All such changes have the potential to influence 
populations, including their risk of extirpation from an area 
[@GarciaCarerras2013; @Vasseuretal2014].
For instance, @frederiksen2008demographic used $43$ years of ring-recovery data 
to examine the demographic impact of ECEs on European shags 
via their effects on temporal variability in 
juvenile, immature and adult survival rates.
@mcdermott2017sensitivity investigated the sensitivity 
of $41$ UK butterfly species to four different types of ECEs. 
At the community level, research has shown that ECEs can reorder
the dominance hierarchy of species in an area, sometimes leading to 
selective local extinctions, and can also produce effects that cascade 
across trophic levels [@thomas2004extinction; @edgar2010nino; @moreno2011extreme; @hoover2014resistance]. 
For example, @chiu2013impact showed that extreme floods affected 
the annual survival of a prey community, in turn, influencing predators in 
aquatic as well as adjacent riparian habitats. 
@dreesen2014successive, in an experimental study, showed that 
successive ECEs with short time intervals between them
can weaken the resistance of plant communities.
<!--***DAN: Shyamolina, what, precisely, does it mean to "weaken the resistance of a
plant community"? Resistance to what? Please clarify.

***Shya: From @dreesen2014successive, authors concluded 
"Our results show that when experimental temperate herbaceous assemblages were exposed to an extreme climatic
event, their resistance to a subsequent event, in terms of leaf
survival, is reduced only when the recovery period is relatively short (14 days for the assemblages studied here) and
not when it is longer."-->

@@

<!--concept for correlation in the extreme tail: tail association-->
Though earlier literature considered the influence of spatial synchrony 
in environmental variables on metapopulation extinction risk, to 
our knowledge the distinct influence of "asymmetric tail associations" between
environmental time series in different locations has hardly been considered. 
"Tail associations" are
associations between two random quantities in their extreme values. 
For instance, if two positively associated 
variables are more associated with their smaller values than in their larger values,
they are said to have stronger
"left-tail" or "lower-tail" association (Figure \ref{fig_intro_cop}A, B). 
If two positively associated variables are more associated in their larger 
values than in their smaller values, they are 
said to have stronger "right-tail" or "upper-tail" association 
(Figure \ref{fig_intro_cop}C, D). 
Figure \ref{fig_intro_cop} shows that two variables can have a wide 
range of patterns of tail association, from extreme left- to extreme right-tail
association, with the same overall correlation. So the tail association
is a distinct concept from correlation itself and may have distinct ecological
ramifications [@Ghosh_copula]. Asymmetric tail associations have been documented 
between environmental variables and between the same environmental variable 
measured in different places [@Serinaldi2008; @li2013; @She2018; @Goswami2018]. 
Asymmetric tail associations have also been documented between numerous ecological
variables and between environmental and ecological variables [@Ghosh_copula].

<!--ECEs probably produce tail associations-->
There are reasons to believe that ECEs may produce asymmetric tail 
associations between climatic variables
measured in different locations. For instance, if a single ECE is associated
with spatially widespread, large values 
of some climatic variable, but small values of the
variable are associated with local weather rather than ECEs, the result could
be a stronger right-tail association between measurements of the climatic variable
made in different locations, through time.
For example, rainstorms which deposit exceptionally large amounts of rain
may do so over large areas, but months with less rain may experience
scattered showers only, which may be much less correlated across locales.

@@

There are also reasons to hypothesize that asymmetry of tail association in 
environmental variables may influence extinction risks for metapopulations.
For instance, assuming for the sake of a simple example that low values of an environmental
variable are "bad" for populations of a given species and high values are "good",
greater left-tail associations between measurements of the variable in different
habitat patches may cause higher metapopulation extinction risk because then
very bad years for component populations would occur at the same time in 
many patches, reducing the potential for rescue effects. Greater left-tail associations
would probably not have the same effect, even if overall correlations between locations
were the same, because it that case it would be very good years for the component 
populations that would occur at the same time in many patches [@Ghosh_copula]. 
@Ghosh_copula substantiated this intuition by showing that it held true for a 
modeling setup using a spatial version of a very simple non-density dependent
population growth model (the Lewontin-Cohen model). In essence, if environmental
"catastrophes" (i.e., extremely bad years for a population) are widely 
spatially synchronized it should create much greater metapopulation extinction 
risk than if "bonanzas" (i.e., extremely good years) are widely synchronized, even
if overall (non-tail-specific) levels of environmental synchrony are the same
in both these cases. The present study extends tests of this idea to metapopulation
models incorporating density dependence.

@@

<!--objectives, last para of intro-->
In this paper, we will explore the potential for asymmetric tail associations to 
influence extinction risk in populations and metapopulations. Specifically,
we will address the following sets of questions. Q1) Can the nature of tail associations
between vital rates in an age-structured population model influence
extinction risk for the model, and to what extent? Q2) Does the nature
of tail associations between measurements of an environmental variable in different
locations influence the extinction risk of a density-dependent metapopulation model
that depends on the environmental variable? How does the nature of the density
dependence mediate this influence? Will density dependence reverse or reproduce
the earlier result [@Ghosh_copula], using non-density-dependent models, that 
left-tail dependence in environmental variables accentuates and right-tail dependence 
mitigates extinction risks?
For Q1, we use a simple 2-stage model. We hypothesize that left-tail dependence
between juvenile survival and adult fecundity rates increases extinction risk
relative to right-tail dependence, even when the overall correlation is kept fixed. 
For Q2, we consider spatial extensions of several simple and commonly
used population models. Though tail associations are used
to assess risks in fields such as finance and hydrology 
[@alexander2008market; @li2013; @chen2011flood; @Goswami2018], to our
knowledge asymmetric tail associations have rarely been assessed for their
importance for extinction risks of populations and species (but see @Ghosh_copula). 
Our questions are an important step for understanding
the potential influence of ECEs on extinction risks if ECEs are associated with
tail dependence in environmental variables.

@@

<!--FIG : pedagogical fig showing 4 types of tail associations: 
extreme left, moderate left, moderate right, extreme right-->
\begin{figure*}[!h]
\begin{center}
\includegraphics[width=15 cm]{./Results/introfig_copula.pdf}
\caption{Two stochastic quantities can have same Kendall 
correlation, but different degrees of tail association: 
(A) extreme left-tail association, (B) moderate left-tail association, (C) 
moderate right-tail association, and (D) extreme right-tail association.
In all cases, Kendall's $\tau$ was the same ($\tau \approx$ `r ktau_common`),
up to sample variation (the sample correlation for each dataset is displayed
on its panel). The extreme left-tail (respectively, right-tail) dependent 
case is perfectly correlated below (respectively, above) the threshold of 0.
See Methods for information on how bivariate noise was generated with
asymmetric tail associations.
\label{fig_intro_cop}}
\end{center}
\end{figure*}
<!--***DAN: Shyamolina, can you please put the actual Kendall tau that pertained
for each dataset on each corresponding
panel? Use, say, 3-4 digits accuracy for display. These values should be very similar, 
but are expected to differ due to sampling variation. I would like to make that
clear.
***Shya: correlation added (rounded upto 4 digit) in each plot.-->

@@

<!--stochastic matrix model : age-structured-->
To investigate (Q1) from \nameref{Introduction} that how do vital rates with an asymmetric tail 
association affect the extinction risk of
populations in a single patch age-structured model,
we considered a simple stochastic age-structured model (Eq. \ref{eq_age_str}) 
with two age classes (egg $E$ and female adult $A$).

\begin{equation}
\left({\begin{array}{cc} E \\ A \end{array}}\right)_{t+1} = 
\left(\begin{array}{cc} 0 & f_a\\ s_e & 0 \end{array}\right)
\left(\begin{array}{c} E \\ A \end{array}\right)_{t} \label{eq_age_str}
\end{equation}

\begin{equation}
f_a \sim \Gamma (k, \theta)\exp(-A/\Omega), \hspace{0.2 cm} s_e \sim \mathcal{B}(\alpha,\beta) \label{eq_vitalrates}
\end{equation}

We assumed the population was semelparous, i.e., adult ($A$) went through the
reproductive episode only once before death. Eggs and adults in the 
next generation evolved depending on the fecundity of adult ($f_a \geq 0$) 
and survival of eggs ($0 \leq s_e \leq 1$) from present generation.
Fluctuating vital rates ($f_a$ and $s_e$) captured the environmental
stochasticity (Eq. \ref{eq_vitalrates}). We made adult's fecundity 
($f_a$) negative density-dependent with its own population. When $\Omega = \infty$ in 
Eq. \ref{eq_vitalrates}, the projection matrix was independent of populations 
(i.e., Eq. \ref{eq_age_str} was "linear" model then), 
but when $\Omega$ was finite, Eq. \ref{eq_age_str} became "nonlinear" with density dependence.
Fecundity ($f_a$) and survival ($s_e$) rates were random draw from a gamma ($\Gamma(k,\theta)$) and
a beta ($\mathcal{B}(\alpha,\beta)$) distribution, respectively. We generated those distributions 
from iid noises with a varied degree of asymmetric tail association between 
fecundity ($f_a$) and survival ($s_e$) rates. Degree of tail association had two broad 
categories: moderate and extreme. Each category had specific direction in tail association: 
either in left or in right tail of vital rates (as depicted in Fig. \ref{fig_intro_cop}). Vital 
rates ($f_a$ and $s_e$) were left tail dependent when unusual bad years ("catastrophes") 
suppressed both vital rates synchronously, 
and right tail dependent when unexpected good years ("bonanzas") boosted both of them.

The moderate tail association was modeled by using the 
concept of "copula", a statistical tool to model 
dependencies. Appendix \ref{SM-intro_copula} summarized
a brief introduction of "copulas" and their relation to tail associations, symmetries 
with examples of available copula families 
from `R` packages (`copula` and `VineCopula`). We used bivariate Clayton 
copula to model moderate left-tail and Survival Clayton copula to model 
moderate right-tail association in the vital rates. Both copulae had the same Kendall 
correlation ($\tau =$ `r ktau_common`) but opposite association in their respective tails. 
We generated a $N \times 2$ matrix, M, for each of the left (Clayton) and 
right (Survival Clayton) tail associated bivariate copulas (see `BiCopSim` function from
`VineCopula` package ), where $N$ is number of years ($=25$) multiplied 
by number of simulations ($=10000$) we used. 
We chose parameters for each copula family using 
`iTau` function from `copula` package so that both copula had
same Kendall's correlation ($\tau =$ `r ktau_common`). Entries within each column of
matrix $M$, then became independent of each other but 
entries between columns had a specific tail association
with positive Kendall's correlation.

On a similar note, we generated extreme left- (or right-) tail dependent matrix (M)
from standard iid (Appendix \ref{SM-extr_tail_dep}) and made the marginals uniform so that
it could mimic an extreme tail dependent copula. We made the first column of matrix 
$M$ as gamma ($\Gamma(k,\theta)$) distributed 
with shape parameter $k = 2$ and scale parameter $\theta = 1$.
This distribution represented fecundity of the adult population, $f_a \geq 0$.
For a given $\theta$, increasing $k$ would shift the distribution towards
higher fecundity and reduce the extinction risk if the other parameters
remain unchanged. We took the second column of $M$ and applied a 
beta ($\mathcal{B}(\alpha,\beta)$) transformation. $\alpha =3$ and
$\beta = 2$ were the shape parameters that could control the distribution 
of the survival rates of eggs ($s_e$) ranging in between 0 to 1. 
For each $25$ timesteps from each of $10000$ simulations,
we incorporated pairwise entries from each row of matrix $M$ in the expression for 
vital rates (fecundity $f_a$ and survival rates $s_e$, Eq. \ref{eq_vitalrates}) of 
the stochastic projrction matrix in model (\ref{eq_age_str}).

<!--calculation of ext risk with age-structured model-->
To calculate extinction risk after each time step, we simulated model (\ref{eq_age_str}) for $25$ years
with $10000$ simulations. For each simulation, we started with initial population $(E,A)_{t=0} = (5,10)$
and following Eq. \ref{eq_age_str} we checked whether after each time step the total population $(E+A)_{t}$
was below a certain extinction threshold (we set it as $2$) or not. If total population was below that threshold then we 
considered population went extinct for that simulation. We also calculated $2.5^{th}$ and $97.5^{th}$ quantiles
of $10000$ simulations with the probability of computed extinction risk at each time step (see [`code`](https://github.com/sghosh89/ERC/blob/master/ext_risk_cop_age_str.R)).

>>>>

---
title: "Tail associations and their impact on extinction risk"
author: "Shyamolina Ghosh, Daniel Reuman"
date: "`r format(Sys.time(), '%d %B, %Y')`"
geometry: "left=1in,right=1in,top=1in,bottom=1in"
output: 
  pdf_document:
    number_sections: no
    keep_tex: yes
    fig_caption: yes
    includes:
      in_header: head_maintext.sty

@@

<!--
Alternative titles
1) Tail associations in environmental variables and extinction risk
2) The influence of correlations in the extremes on extinction risk
3) Tail associations in environmental variables and their impact on extinction risk
4) Asymmetric tail association in extreme climatic events affects population-extinction risk
-->

@@

<!--ECEs have a lot of impacts, it is important to understand their impacts-->
Over the past few decades, extreme climatic events (ECEs) such as
temperature or precipitation extremes, heat-waves, and extreme wildfires
have become more common or more extreme due to anthropogenic climate change 
[@ummenhofer2017extreme]. ECEs are well known to produce severe
impacts at the population, community and ecosystem levels [@bragazza2008climatic; 
@gutschick2003extreme; @jiguet2011community; @mckechnie2009climate;
@felton2017integrating], and thus the goal of improving our understanding of
the variety of ecological impacts of ECEs is an important applied goal for 
modern ecology [@katz2005statistics; @smith2011ecological; @diez2012will; @moritz2013future; @bailey2016tackling]. Frequent extreme events 
can help alter the mean temperature or precipitation of an area, 
can be associated with changes through time in the overall variance of 
climate signals, and can directly contribute to changes in the symmetry or the 
skewness of distributions of climate signals through time 
[@hansen2012perception]. All such changes have the potential to influence 
populations, including their risk of extirpation from an area 
[@GarciaCarerras2013; @Vasseuretal2014].
For instance, @frederiksen2008demographic used $43$ years of ring-recovery data 
to examine the demographic impact of ECEs on European shags 
via their effects on temporal variability in 
juvenile, immature and adult survival rates.
@mcdermott2017sensitivity investigated the sensitivity 
of $41$ UK butterfly species to four different types of ECEs. 
At the community level, research has shown that ECEs can reorder
the dominance hierarchy of species in an area, sometimes leading to 
selective local extinctions, and can also produce effects that cascade 
across trophic levels [@thomas2004extinction; @edgar2010nino; @moreno2011extreme; @hoover2014resistance]. 
For example, @chiu2013impact showed that extreme floods affected 
the annual survival of a prey community, in turn influencing predators in 
aquatic as well as adjacent riparian habitats. 
@dreesen2014successive, in an experimental study, showed that 
ECEs can reduce the ability of plant communities to withstand subsequent ECEs
that occur after a short time interval.

@@

<!--concept for correlation in the extreme tail: tail association-->
Though earlier literature considered the influence of spatial synchrony 
in environmental variables on metapopulation extinction risk, to 
our knowledge the distinct influence of "asymmetric tail associations" between
environmental time series in different locations has hardly been considered. 
"Tail associations" are
associations between two random quantities in their extreme values. 
For instance, if two positively associated 
variables are more closely associated in their smaller values than in their larger values,
they are said to have stronger
"left-tail" or "lower-tail" association (Figure \ref{fig_intro_cop}A, B). 
If two positively associated variables are more associated in their larger 
values than in their smaller values, they are 
said to have stronger "right-tail" or "upper-tail" association 
(Figure \ref{fig_intro_cop}C, D). 
Figure \ref{fig_intro_cop} shows that two variables can have a wide 
range of patterns of tail association, from extreme left- to extreme right-tail
association, with the same overall correlation. So tail association
is a distinct concept from correlation itself, and may have distinct ecological
ramifications [@Ghosh_copula]. Asymmetric tail associations have been documented 
between environmental variables, and between the same environmental variable 
measured in different places [@Serinaldi2008; @li2013; @She2018; @Goswami2018]. 
Asymmetric tail associations have also been documented between numerous ecological
variables and between environmental and ecological variables [@Ghosh_copula].

<!--ECEs probably produce tail associations-->
There are reasons to believe that ECEs may produce asymmetric tail 
associations between climatic variables
measured in different locations. For instance, if a single ECE is associated
with spatially widespread, large values 
of some climatic variable, but small values of the
variable are associated with local weather rather than ECEs, the result could
be stronger right-tail association between measurements of the climatic variable
made in different locations, through time.
For example, rainstorms which deposit exceptionally large amounts of rain
may do so over large areas; but months with less rain may experience
scattered showers only, which may be much less correlated across locales.

@@

There are also reasons to hypothesize that asymmetry of tail association in 
environmental variables may influence extinction risks for metapopulations.
For instance, assuming for the sake of a simple example that low values of an environmental
variable are "bad" for populations of a given species and high values are "good",
greater left-tail associations between measurements of the variable in different
habitat patches may cause higher metapopulation extinction risk because then
very bad years for component populations would occur at the same time in 
many patches, reducing the potential for rescue effects. Greater left-tail associations
would probably not have the same effect, even if overall correlations between locations
were the same, because in that case it would be very good years for the component 
populations that would occur at the same time in many patches [@Ghosh_copula]. 
@Ghosh_copula substantiated this intuition by showing that it held true for a 
modeling setup using a spatial version of a very simple non-density dependent
population growth model (the Lewontin-Cohen model). In essence, if environmental
"catastrophes" (i.e., extremely bad years for a population) are widely 
spatially synchronized it should create much greater metapopulation extinction 
risk than if "bonanzas" (i.e., extremely good years) are widely synchronized, even
if overall (non-tail-specific) levels of environmental synchrony are the same
in both these cases. The present study extends tests of this idea to metapopulation
models incorporating density dependence.

@@

<!--objectives, last para of intro-->
In this paper, we will explore the potential for asymmetric tail associations to 
influence extinction risk in populations and metapopulations. Specifically,
we will address the following sets of questions. Q1) Can the nature of tail associations
between vital rates in an age-structured population model influence
extinction risk for the model, and to what extent? Q2) Does the nature
of tail associations between measurements of an environmental variable in different
locations influence the extinction risk of a density-dependent metapopulation model
that depends on the environmental variable? How does the nature of the density
dependence mediate this influence? Will density dependence reverse or reproduce
the earlier result [@Ghosh_copula], using non-density-dependent models, that 
left-tail dependence in environmental variables accentuates and right-tail dependence 
mitigates extinction risks?
For Q1, we simulate a simple 2-stage model. We hypothesize that left-tail dependence
between juvenile survival and adult fecundity rates increases extinction risk
relative to right-tail dependence, even when the overall correlation is kept fixed. 
For Q2, we simulate spatial extensions of several simple and commonly
used population models. We intentionally use a simulation approach and simple models
instead of attampting mathematical analysis and/or using realistic models
because our intention here is to provide an initial exploration only of 
the potential mechanisms outlined above which suggest that
asymmetries of tail association may make a difference for extinction risks. 
<!--***DAN: In Discussion need to talk about what analytic results are needed
to take the next step of determining when tail depencence may make enough of
a difference to think about. I think our results, together with Tulja's 
approximation, suggest that tail associations probably don't make much difference
for models without a spatial component. But our results illsutaret that it can
make a big difference when there is a spatial component, apparently because of 
the influence of tail associations on rescue effects. Could maybe even outline
analytic approaches for non-dd models? Not sure what would be done with dd
models. The point is you need to take the next step in the Discussion of 
talking about how our results, which admitedly are just for specific models,
could be turned into analytic results or general results.-->
Though tail associations are used
to assess risks in fields such as finance and hydrology 
[@alexander2008market; @li2013; @chen2011flood; @Goswami2018], to our
knowledge asymmetric tail associations have rarely been assessed for their
importance for extinction risks of populations and species (but see @Ghosh_copula). 
Our questions are an important step for understanding
the potential influence of ECEs on extinction risks if ECEs are associated with
tail dependence in environmental variables.

@@

<!--FIG : pedagogical fig showing 4 types of tail associations: 
extreme left, moderate left, moderate right, extreme right-->
\begin{figure*}[!h]
\begin{center}
\includegraphics[width=15 cm]{./Results/introfig_copula.pdf}
\caption{Two stochastic quantities can have same Kendall 
correlation, but different degrees of tail association: 
(A) extreme left-tail association, (B) moderate left-tail association, (C) 
moderate right-tail association, and (D) extreme right-tail association.
In all cases, Kendall's $\tau$ was the same, $\tau = 0.75$,
up to sample variation (the sample correlation for each dataset is displayed
on its panel). The extreme left-tail (respectively, right-tail) dependent 
case is perfectly correlated below (respectively, above) the threshold of 0.
See Methods for information on how bivariate noise was generated with
asymmetric tail associations.
\label{fig_intro_cop}}
\end{center}
\end{figure*}
<!--***DAN: Shyamolina, it occured to be there are simple and good analytic reasons
why ktau_common is exactly 0.75 (up to sample variation). 
I bet you will see them too if you look up the
definition of kendall tau and think about the extreme tail dependent cases. 
Currently, around line 50 of the sup mat Rmd, you calculate ktau_common
by simulating 100000 points from the extreme-tail-denepdnece noise and then
calculating the sample kendall. Instead, can you please replace those lines 
with the simple line "ktau_common<-0.75"? I would do this myself but I
don't want to risk messing up your code in some way I don't understand. You should not
need the seed anymore to ensure you get the same ktau_common on each run
since you won't be using the sample correlartion anymore to estimate ktau_common,
you'll just be using the analytic result that the population correlation
of data from the extreme-tail-dependent algorithm (in the limit of infinite
sample sizes) is exactly 0.75. Please
let me know if you do not understand something here.-->
<!--***DAN: Shyamolina, currently the sample kendalls are the same on A and D, and
also on B and C, and I know why. This looks weird, and makes the reader
wonder why this is the case. Rather than plotting the reversed version of the
data from A on D, and the reversed version of the data from B on C, can you plase
generate separate, independent data for A and D, and also for B and C? That
will have the effect of making the sample kendalls on the 4 panels slightly
different, which will mean readers think anything funny is going on.-->

@@

<!--stochastic matrix model : age-structured-->
To investigate (Q1) from the \nameref{Introduction},
we considered a simple stochastic age-structured model
with two age classes,  
\begin{equation}
\left({\begin{array}{cc} E(t+1) \\ A(t+1) \end{array}}\right) = 
\left(\begin{array}{cc} 0 & f_a(t)\\ s_e(t) & 0 \end{array}\right)
\left(\begin{array}{c} E(t) \\ A(t) \end{array}\right), \label{eq_age_str}
\end{equation}
\noindent where $E(t)$ and $A(t)$ are the numbers or population densities of eggs 
and adults, respectively, at time $t$. We chose a very simple model
because our goal was to carry out an initial evaluation, only, of 
the realism of a potential mechanism. The stochastic vital rates $s_e(t)$ and $f_a(t)$
were selected, independently across times, as follows. 
The egg survival rate $s_e(t)$ was drawn from a beta distribution with parameters
$\alpha=3$ and $\beta=2$. 
Adult fecundity was $f_a(t)=\tilde{f}_a(t)\exp(-A(t)/\Omega)$, where $\tilde{f}_a(t)$
was drawn from a gamma distribution with shape and scale parameters $k=2$
and $\theta=1$, respectively, 
and $\Omega$ is a density dependence parameter. Taking $\Omega=\infty$ made
$\exp(-A(t)/\Omega)=1$ so that the model did not include density dependence
(it was then a standard stochastic matrix model). Finite, positive values of 
$\Omega$ made the model nonlinear, with density dependence in the adult
fecundity rate.

The stochastic quantities $s_e(t)$ and $\tilde{f}_a(t)$ were generated 
to have the same overall association (Kendall's $\tau$ equal to $0.75$,
up to sampling variation), but different degrees and types of 
asymmetry of tail association in different simulations, as follows.
For a simulation of length $T$, we used methods that can 
generate $T$ independent pairs $(u(t),v(t))$ for $t=1,\ldots,T$ with
$\tau=0.75$ (up to sample variation), but with either extreme left-,
moderate left-, moderate right-, or extreme right-tail association 
(e.g., Fig. \ref{fig_intro_cop}). The methods we used generated
$(u(t),v(t))$ such that $u(t)$ was uniformly distributed on $(0,1)$,
and likewise for $v(t)$. We then applied a monotonic transformation
to the $u(t)$ so that the resulting numbers were distributed according to
a beta distribution with parameters
$\alpha=3$ and $\beta=2$ 
and used those values for the $s_e(t)$. We also applied a (different)
monotonic transformation to the $v(t)$ so that the resulting numbers were
distributed according to a gamma distribution with parameters $k=2$
and $\theta=1$ and used those values for the $\tilde{f}_a(t)$. Because
the $u(t)$ and $v(t)$ were uniformly distributed, the inverse 
cummulative distribution functions of the desired beta and
gamma distributions were the transformations. Because these
were monotonic transformations, the Kendall correlation of 
$s_e(t)$ and $\tilde{f}_a(t)$ was the same as that of 
$u(t)$ and $v(t)$. The quantities $(u(t),v(t))$ for $t=1,\ldots,T$ were
generated as described in Appendix \ref{SM-makenoise}.

<!--calculation of ext risk with age-structured model-->
To calculate extinction risk for a given $\Omega$ and choice of tail dependence, 
we simulated the model for $25$ time steps, $10000$ times,
each simulation starting from $E(t)=5$ and $A(t)=10$. The population for a simulation
was considered to have gone extinct if the total population $E(t)+A(t)$ ever 
went below an extinction threshold (we used a threshold of $2$) within the 
$25$-time-step time horizon. Extinction risk was the fraction of $10000$
simulations that went extinct.

<!--***DAN: Shyamolina, the below text makes me wonder if I am misunderstanding something, please
explain. The part that confuses me is the 2.5th and 97.5th quantile part, what is the meaning
of that and what does it have to do with extinction riks?

after each time step, we simulated model (\ref{eq_age_str}) for $25$ years
with $10000$ simulations. For each simulation, we started with initial population $(E,A)_{t=0} = (5,10)$
and following Eq. \ref{eq_age_str} we checked whether after each time step the total population $(E+A)_{t}$
was below a certain extinction threshold (we set it as $2$) or not. If total population was below that threshold then we 
considered population went extinct for that simulation. We also calculated $2.5^{th}$ and $97.5^{th}$ quantiles
of $10000$ simulations with the probability of computed extinction risk at each time step (see [`code`](https://github.com/sghosh89/ERC/blob/master/ext_risk_cop_age_str.R)).
-->

### 2019-10-25 SuppMat.Rmd
working on methods, also some other points made

---
title: "Supporting Information : Asymmetric tail association in extreme climatic events affects population-extinction risk"
author: "Shyamolina Ghosh, Daniel Reuman"
date: "`r format(Sys.time(), '%d %B, %Y')`"
geometry: "left=1in,right=1in,top=1in,bottom=1in"

@@

# A brief introduction about "copulas"  \label{intro_copula}

<!--needs editing, I just dumped old text here-->
Copula, a rank based measure of association, is much more informative 
than linear correlation measure and a widely applied tool in dependence 
modeling [@joe2014_dependence; @nelsen2006_copula]. Copulas separate 
the information content of a bivariate dataset,
$(x_n, y_n)$ for $n = 1 \ldots N$, into two non-overlapping parts: the information in the marginal 
distributions (which is not about the association between the variables) and the rest of 
the information (which is solely about the association). The association information is revealed by 
the copula plot ($v_n$ against $u_n$) of the raw variables ($y_n$ vs. $x_n$) in Fig...., where 
$u_n$ is the rank of $x_n$ in the set ${x_1, x_2, \ldots , x_n }$, divided by $N + 1$; 
and $v_n$ is the rank of $y_n$ in the set ${y_1, y_2, \ldots , y_n }$, also divided
by $N + 1$. Here the rank of the smallest element of each sets is considered to be 1 and the rank of
the largest element is $N$. We refer to the $u_n$ and $v_n$ as normalized ranks of the $x_n$ and $y_n$. 
Ranking makes the marginal distributions uniform, isolating only the information on 
association between the variables.

# Algorithm to generate extreme tail association  \label{extr_tail_dep}

@@

>>>>

---
title: "Supporting information: Tail associations and their impact on extinction risk"
author: "Shyamolina Ghosh, Daniel Reuman"
date: "`r format(Sys.time(), '%d %B, %Y')`"
geometry: "left=1in,right=1in,top=1in,bottom=1in"

@@

# Generating noise with asymmetric tail associations  \label{makenoise}

<!--needs editing, I just dumped old text here-->

Copula, a rank based measure of association, is much more informative 
than linear correlation measure and a widely applied tool in dependence 
modeling [@joe2014_dependence; @nelsen2006_copula]. Copulas separate 
the information content of a bivariate dataset,
$(x_n, y_n)$ for $n = 1 \ldots N$, into two non-overlapping parts: the information in the marginal 
distributions (which is not about the association between the variables) and the rest of 
the information (which is solely about the association). The association information is revealed by 
the copula plot ($v_n$ against $u_n$) of the raw variables ($y_n$ vs. $x_n$) in Fig...., where 
$u_n$ is the rank of $x_n$ in the set ${x_1, x_2, \ldots , x_n }$, divided by $N + 1$; 
and $v_n$ is the rank of $y_n$ in the set ${y_1, y_2, \ldots , y_n }$, also divided
by $N + 1$. Here the rank of the smallest element of each sets is considered to be 1 and the rank of
the largest element is $N$. We refer to the $u_n$ and $v_n$ as normalized ranks of the $x_n$ and $y_n$. 
Ranking makes the marginal distributions uniform, isolating only the information on 
association between the variables.

<!--# Algorithm to generate extreme tail association  \label{extr_tail_dep}-->

@@

<!--some stuff that used to be in the main text-->

The moderate tail association was modeled by using the 
concept of "copula", a statistical tool to model 
dependencies. Appendix \ref{SM-intro_copula} summarized
a brief introduction of "copulas" and their relation to tail associations, symmetries 
with examples of available copula families 
from `R` packages (`copula` and `VineCopula`). We used bivariate Clayton 
copula to model moderate left-tail and Survival Clayton copula to model 
moderate right-tail association in the vital rates. Both copulae had the same Kendall 
correlation ($\tau =$ `r ktau_common`) but opposite association in their respective tails. 
We generated a $N \times 2$ matrix, M, for each of the left (Clayton) and 
right (Survival Clayton) tail associated bivariate copulas (see `BiCopSim` function from
`VineCopula` package ), where $N$ is number of years ($=25$) multiplied 
by number of simulations ($=10000$) we used. 
We chose parameters for each copula family using 
`iTau` function from `copula` package so that both copula had
same Kendall's correlation ($\tau =$ `r ktau_common`). Entries within each column of
matrix $M$, then became independent of each other but 
entries between columns had a specific tail association
with positive Kendall's correlation.

<!--some more stuff from the main text-->

On a similar note, we generated extreme left- (or right-) tail dependent matrix (M)
from standard iid (Appendix \ref{SM-extr_tail_dep}) and made the marginals uniform so that
it could mimic an extreme tail dependent copula. We made the first column of matrix 
$M$ as gamma ($\Gamma(k,\theta)$) distributed 
with shape parameter $k = 2$ and scale parameter $\theta = 1$.
This distribution represented fecundity of the adult population, $f_a \geq 0$.
For a given $\theta$, increasing $k$ would shift the distribution towards
higher fecundity and reduce the extinction risk if the other parameters
remain unchanged. We took the second column of $M$ and applied a 
beta ($\mathcal{B}(\alpha,\beta)$) transformation. $\alpha =3$ and
$\beta = 2$ were the shape parameters that could control the distribution 
of the survival rates of eggs ($s_e$) ranging in between 0 to 1. 
For each $25$ timesteps from each of $10000$ simulations,
we incorporated pairwise entries from each row of matrix $M$ in the expression for 
vital rates (fecundity $f_a$ and survival rates $s_e$, Eq. \ref{eq_vitalrates}) of 
the stochastic projrction matrix in model (\ref{eq_age_str}).
